Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
add examples explanations

It is often necessary to relate certain invariants in the pre- and
postcondition when specifying a program.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving
triples in the logic, so the Coq theorem for the above example would
look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function $S_f$ body with this
induction hypothesis.  In this process it can be the case that the
induction hypothesis needs to be used with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

@@

@@

\begin{figure}
\begin{lstlisting}
${\color{blue} \{ A(0, N) + N \} }$
i=1; m=a[0];
${\color{blue}%
  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-1) \} }$
while (i < N) {
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
  if (a[i] > m)
    m=a[i];
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
  i=i+1;
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
} ${\color{blue} \{ 0 \} }$
\end{lstlisting}
\caption{Example derivation where we noted $A(i, N)$
  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
  the metric used here assigns a cost of 1 to every assignment
  and 0 to all other operations.
  }
\label{fig:xmplmax}
\end{figure}

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ (Z = \log_2(h-l)) + Z \} }$
bsearch(x,l,h) {
  if (h-l > 1) {
    ${\color{blue}%
      \{ (h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
    m = h + (h-l)/2;
    ${\color{blue}%
      \{ (m = \frac{h+l}{2} \land h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
    if (a[m]>x) h=m; else l=m;
    ${\color{blue}%
      \{ (Z - 1 = \log_2(h-l)) + (Z-1) + 1 \} }$
    l = bsearch(x,l,h); }
  ${\color{blue}%
    \{ Z-1 \} }$
  return l;
} ${\color{blue} \{ Z \} }$
\end{lstlisting}
\caption{Example derivation of a stack usage bound for a binary
  search program, the metric used assigns a cost of 1 to function
  calls and $-1$ to return statements, we noted $Z$ an arbitrary
  logical variable.
  }
\end{figure}

>>>>

It is often necessary to relate certain invariants in the pre- and
postcondition when specifying a program.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving
triples in the logic, so the Coq theorem for the above example would
look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function $S_f$ body with this
induction hypothesis.  In this process it can be the case that the
induction hypothesis needs to be used with a different value of a
logical variable.  To cope with this problem, assumptions in the
function context $\Delta$ are universally quantified over logical
variables.  The {\sc L:Extend} rule uses the host proof assistant to
require that the triple on $f$'s body is proved for every possible
logical variable $y$.

@@

\paragraph{Using the Quantitative Logic.}

Our first example on~\pref{fig:xmplmax} uses the expressive power of
the logic to get a very precise runtime bound on a program searching
a maximal element in an array.  The metric used simply counts the
assignments performed by the program.  Hence, the resource consumption
is closely related to the number $A$ of times the test \lstinline{a[i] > m}
is true during the execution.
%
It is clear that $A+1$ is the count of ``maximums so far in the
array \lstinline{a}'':
$$
  \# \{ k \mid 0 \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \},
$$
where we write $\# S$ for the cardinal of the set $S$.  This number is
obviously bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But using the expressivity of our logic we can get a finer bound
that depends on the contents of the array before the program execution.

The only non-trivial part of this derivation is finding the loop
invariant that is used in the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program; it can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
is admissible in our system.

The second example on~\pref{fig:xmplbs} shows a use case for
logical variables as well as a metric for stack consumption.
We are interested in showing that a binary search function
$\code{bsearch}$ has logarithmic stack consumption.  We use
a logical variable $Z$ in the function specification
$\htriple{(Z = \log_2(h-l)) + Z \cdot F}{}{Z \cdot F}$
to express that the stack required by the function is fully
returned afterwards.

The critical step in this proof is the application of the
{\sc L:Call} rule to prove the recursive call.  At this point
the context $\Delta$ contains the following specification
for the function \code{bsearch}
$$
  \forall y \, (x,l,h) \, \_.
  ( (\lambda y \, (\_, l, h) .\, (y {=} \log_2(h {-} l)) + y {\cdot} F)
  , (\lambda y \, \_ .\, y {\cdot} F)
  ).
$$
Using the {\sc L:Call} rule it is possible to instantiate
$y$ with $Z - 1$, and because $M_f = F$ and $M_r = -F$ in
the stack resource metric, the appropriate triple can be
derived on the call site.  The rest of the proof does not
involve any resource manipulation and is just bookkeeping
of logical assertions.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue} \{ A(0) + N \} }$
i=1; m=a[0];
${\color{blue}%
  \{ (i {=} 1 \land m=a[0]) + A(1) + (N-1) \} }$
while (i < N) {
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
  if (a[i] > m)
    m=a[i];
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1) + (N-i) \} }$
  i=i+1;
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i) \} }$
} ${\color{blue} \{ 0 \} }$
\end{lstlisting}
\caption{Example derivation where we wrote $A(i)$
  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$,
  the metric used here assigns a cost of 1 to every assignment
  and 0 to all other operations.
  }
\label{fig:xmplmax}
\end{figure}

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
bsearch(x,l,h) {
  if (h-l > 1) {
    ${\color{blue}%
      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
    m = h + (h-l)/2;
    ${\color{blue}%
      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
    if (a[m]>x) h=m; else l=m;
    ${\color{blue}%
      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
    l = bsearch(x,l,h); }
  ${\color{blue}%
    \{ (Z-1) \cdot F - (-F) \} }$
  return l;
} ${\color{blue} \{ Z \cdot F \} }$
\end{lstlisting}
\caption{Example derivation of a stack usage bound for a binary
  search program.  The metric used sets a cost of $F$ before
  to function call and $-F$ after the call, $F$ is the stack
  frame size of the function $\code{bsearch}$.  We use $Z$ as
  logical variable.
  }
\label{fig:xmplbs}
\end{figure}
